---
title: "Assignment 4: Data Wrangling"
author: "Melissa Merritt"
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Wrangling 

## Directions
1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

The completed exercise is due on Friday, Oct7th @ 5:00pm.

## Set up your session 
1. Check your working directory, load the `tidyverse` and `lubridate` packages, and upload all four raw data files associated with the EPA Air dataset, being sure to set string columns to be read in a factors. See the README file for the EPA air datasets for more information (especially if you have not worked with air quality data previously). 

2. Explore the dimensions, column names, and structure of the datasets. 
```{r setup, message=FALSE, warning=FALSE}

install.packages('formatR')
library(formatR)

knitr::opts_chunk$set(tidy.opts=list(width.cutoff=45), tidy = TRUE)
```

```{r, message=FALSE}
#1 
getwd()

# getwd() allows me to check my working directory. I needed to change it in the knit options to project directory, and now it is correct.

library(tidyverse)
library(lubridate)

O3_NC2018 <- read.csv("./Data/Raw/EPAair_O3_NC2018_raw.csv", stringsAsFactors = TRUE)
O3_NC2019 <- read.csv("./Data/Raw/EPAair_O3_NC2019_raw.csv", stringsAsFactors = TRUE)
PM25_NC2018 <- read.csv("./Data/Raw/EPAair_PM25_NC2018_raw.csv", stringsAsFactors = TRUE)
PM25_NC2019 <- read.csv("./Data/Raw/EPAair_PM25_NC2019_raw.csv", stringsAsFactors = TRUE)

# While importing the data files, I am able to name them in a way that makes them easy to use and recognize.


#2
dim(O3_NC2018)
dim(O3_NC2019)
dim(PM25_NC2018)
dim(PM25_NC2019)

# dim() allows me to check the columns and rows for each data set.

colnames(O3_NC2018)
colnames(O3_NC2019)
colnames(PM25_NC2018)
colnames(PM25_NC2019)

# the colnames() function allows me to see the names of each column in the data set. 

str(O3_NC2018)
str(O3_NC2019)
str(PM25_NC2018)
str(PM25_NC2019)

# The str() function allows me to check the structure of a data set. 

```

## Wrangle individual datasets to create processed files.
3. Change date to date
4. Select the following columns: Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE
5. For the PM2.5 datasets, fill all cells in AQS_PARAMETER_DESC with "PM2.5" (all cells in this column should be identical).
6. Save all four processed datasets in the Processed folder. Use the same file names as the raw files but replace "raw" with "processed".

```{r}
#3
O3_NC2018$Date <- as.Date(O3_NC2018$Date, format = "%m/%d/%Y")
O3_NC2019$Date <- as.Date(O3_NC2019$Date, format = "%m/%d/%Y")
PM25_NC2018$Date <- as.Date(PM25_NC2018$Date, format = "%m/%d/%Y")
PM25_NC2019$Date <- as.Date(PM25_NC2019$Date, format = "%m/%d/%Y")

class(O3_NC2018$Date)
class(O3_NC2019$Date)
class(PM25_NC2018$Date)
class(PM25_NC2019$Date)

# I was able to use the as.Date() function to change the date column to "date". I was able to check each of the classifications with the class() function.

#4
O3_NC2018_select <- select(O3_NC2018, Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY:SITE_LONGITUDE)
O3_NC2019_select <- select(O3_NC2019, Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY:SITE_LONGITUDE)
PM25_NC2018_select <- select(PM25_NC2018, Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY:SITE_LONGITUDE)
PM25_NC2019_select <- select(PM25_NC2019, Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY:SITE_LONGITUDE)

# Using the select() function I am able to create a new data file with just the pertinent information. 

#5

PM25_NC2018_select$AQS_PARAMETER_DESC <- "PM2.5"
PM25_NC2019_select$AQS_PARAMETER_DESC <- "PM2.5"

# Using the column name I was able to rename the information in the columns to "PM2.5"

#6
write.csv(O3_NC2018_select, row.names = FALSE, 
          file = "./Data/Processed/EPAair_O3_NC2018_processed.csv")
write.csv(O3_NC2019_select, row.names = FALSE, 
          file = "./Data/Processed/EPAair_O3_NC2019_processed.csv")
write.csv(PM25_NC2018_select, row.names = FALSE, 
          file = "./Data/Processed/EPAair_PM25_NC2018_processed.csv")
write.csv(PM25_NC2019_select, row.names = FALSE, 
          file = "./Data/Processed/EPAair_PM25_NC2019_processed.csv")

# I am able to use the write.csv function to save the processed files in the processed data folder. 

```


## Combine datasets
7. Combine the four datasets with `rbind`. Make sure your column names are identical prior to running this code. 
8. Wrangle your new dataset with a pipe function (%>%) so that it fills the following conditions: 

* Include all sites that the four data frames have in common: "Linville Falls", "Durham Armory", "Leggett", "Hattie Avenue", "Clemmons Middle", "Mendenhall School", "Frying Pan Mountain", "West Johnston Co.", "Garinger High School", "Castle Hayne", "Pitt Agri. Center", "Bryson City", "Millbrook School" (the function `intersect` can figure out common factor levels)
* Some sites have multiple measurements per day. Use the split-apply-combine strategy to generate daily means: group by date, site, aqs parameter, and county. Take the mean of the AQI value, latitude, and longitude.
* Add columns for "Month" and "Year" by parsing your "Date" column (hint: `lubridate` package)
* Hint: the dimensions of this dataset should be 14,752 x 9.

9. Spread your datasets such that AQI values for ozone and PM2.5 are in separate columns. Each location on a specific date should now occupy only one row.
10. Call up the dimensions of your new tidy dataset.
11. Save your processed dataset with the following file name: "EPAair_O3_PM25_NC1819_Processed.csv"

```{r}
#7
EPAair_O3_PM25_NC1819 <- rbind(O3_NC2018_select, O3_NC2019_select, PM25_NC2018_select, PM25_NC2019_select)

# Using the rbind() function allows me to combine the data files because all the row names are the same. 

#8
EPAair_O3_PM25_NC1819 <- 
  EPAair_O3_PM25_NC1819 %>%
  filter(Site.Name == "Linville Falls" | Site.Name == "Durham Armory" | Site.Name == "Leggett" | Site.Name == "Hattie Avenue" | Site.Name == "Clemmons Middle" | Site.Name == "Mendenhall School" | Site.Name == "Frying Pan Mountain" | Site.Name == "West Johnston Co." | Site.Name == "Garinger High School" | Site.Name == "Castle Hayne" | Site.Name == "Pitt Agri. Center" | Site.Name == "Bryson City" | Site.Name == "Millbrook School") %>%
  group_by(Date, Site.Name, AQS_PARAMETER_DESC, COUNTY) %>%
  summarise(Mean_AQI = mean(DAILY_AQI_VALUE), 
            Mean_latitude = mean(SITE_LATITUDE), 
            Mean_longitude = mean(SITE_LONGITUDE)) %>%
  mutate(Month = month(Date), Year = year(Date))

#I used the filter() function to gather just the sites that existed in all four of the data frames. Then I was able to use the group_by() function to clarify that for each row that had the same values in those four columns, we would take the means of the Daily AQI Value, the site latitude, and the site longitude. Lastly, I used the lubridate package, function mutate(), to create 2 new columns, month and year. 

#9
EPAair_O3_PM25_NC1819_spread <- pivot_wider(EPAair_O3_PM25_NC1819, names_from = AQS_PARAMETER_DESC, values_from = Mean_AQI)

# I used the pivot_wider() function to spread the data out. I got the names from the original AQS_PARAMETER_DESC column, and was able to create two new columns (ozone and PM2.5) with the values from the Mean_AQI. 

#10
dim(EPAair_O3_PM25_NC1819_spread)

# Once again the dim() function gives me the columns and the rows in the data frame.

#11

write.csv(EPAair_O3_PM25_NC1819_spread, row.names = FALSE, file =  "./Data/Processed/EPAair_O3_PM25_NC1819_Processed.csv")

```


## Generate summary tables
12. Use the split-apply-combine strategy to generate a summary data frame. Data should be grouped by site, month, and year. Generate the mean AQI values for ozone and PM2.5 for each group. Then, add a pipe to remove instances where a month and year are not available (use the function `drop_na` in your pipe). 

13. Call up the dimensions of the summary dataset.

```{r}
#12a
EPAair_O3_PM25_NC1819_summary <-
  EPAair_O3_PM25_NC1819_spread %>%
  group_by(Site.Name, Month, Year) %>%
  summarise(mean_AQI_ozone = mean(Ozone),
            mean_AQI_PM2.5 = mean(PM2.5))
#12b
EPAair_O3_PM25_NC1819_summary_NA <-
  EPAair_O3_PM25_NC1819_summary %>%
  drop_na(mean_AQI_ozone, mean_AQI_PM2.5)

# To summarize the data into our final data frame I grouped the data with the group_by() function, and then combined the ozone and PM2.5 into mean values for the rows that shared the same Site.Name, Month, and Year. The drop_NA function allows me to get rid of the NA values in both the mean_AQI_ozone column and the mean_AQI_PM2.5 column. 

#13

dim(EPAair_O3_PM25_NC1819_summary_NA)


```

14. Why did we use the function `drop_na` rather than `na.omit`?

> Answer: We used the function 'drop_na' over 'na.omit' because the na.omit function cannot observe a subset of columns. It will always omit all of the NAs. With the drop_na function, we are able to specify that we only want to omit rows that have NAs in the mean ozone column and the mean PM2.5 column. In this case, we get the same results for both, but in other cases it could be useful. 
