---
title: "Assignment 5: Data Visualization"
author: "Melissa Merritt"
output: pdf_document
geometry: margin=2.54cm
---

```{r setup, include=FALSE}
#This code chunk will tidy your knit PDF files, wrapping long code lines
#For it to work, the "formatR" package needs to be installed

install.packages('formatR')
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE)
```

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

The completed exercise is due on Friday, Oct 14th @ 5:00pm.



## Set up your session 
1. Set up your session. Verify your working directory and load the tidyverse, 
lubridate, & cowplot packages. Upload the NTL-LTER processed data files for  
nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy 
[`NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv`] version) and 
the processed data file for the Niwot Ridge litter dataset (use the
[`NEON_NIWO_Litter_mass_trap_Processed.csv`] version). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r}
#1 
getwd()

# I used getwd() to check the working directory. 

install.packages("tidyverse")
install.packages("lubridate")
install.packages("cowplot")

library(tidyverse)
library(lubridate)
library(cowplot)

# Using the install.packages() and library() to get the packages needed to run the specific functions.

Chemistry_Nutrients_PeterPaul <- read.csv("./Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv", stringsAsFactors = TRUE)

NiwotRidge_Litter <- read.csv("./Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv", stringsAsFactors = TRUE)

# I used the read.csv() function to import the data files and I renamed them to something shorter and easier to work with. 

#2 

class(Chemistry_Nutrients_PeterPaul$sampledate)
class(NiwotRidge_Litter$collectDate)

# Initially when I check the class of the date columns in each of the data frames they were factors, so I used the as.Date() function below to change the class to date. 

Chemistry_Nutrients_PeterPaul$sampledate <- as.Date(Chemistry_Nutrients_PeterPaul$sampledate, format = "%Y-%m-%d")
NiwotRidge_Litter$collectDate <- as.Date(NiwotRidge_Litter$collectDate, format = "%Y-%m-%d")

class(Chemistry_Nutrients_PeterPaul$sampledate)
class(NiwotRidge_Litter$collectDate)

# I was able to recheck the class with the class() function to ensure that the date columns were now being read as date formats. 

```
## Define your theme
3. Build a theme and set it as your default theme.

```{r}
#3
mytheme <- theme_bw(base_size = 12) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top") 

# I built my theme using the black and white theme. Base_size represents the base font, so I set that at 12. The theme() function allows me to change the color scheme of the axis text, and keep the legend on top of the plots. 
```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best 
practices for data visualization. Ensure your theme, color palettes, axes, 
and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with 
separate aesthetics for Peter and Paul lakes. Add a line of best fit and 
color it black. Adjust your axes to hide extreme values (hint: change the 
limits using `xlim()` and/or `ylim()`).

```{r}
#4
Phosphorus_Phosphate <- 
  ggplot(Chemistry_Nutrients_PeterPaul, aes(x = po4, y = tp_ug, color = lakename)) +
  geom_point(size = 1) +
  xlim(c = 0, 20) +
  ylim(c = 0, 100) +
  xlab("Total Phosphate (µg)") +
  ylab("Total Phosphorus (µg)") + 
  geom_smooth(method = lm, color = "black", se = F) +
  labs(color = "Lake Name") +
  mytheme 
print(Phosphorus_Phosphate)


# I am using the ggplot() function to assign po4 to x and tp_ug to y to create a plot of phosphorus (tp_ug) BY phosphate (po4). To make the colors by lake, I set color = lakename. I am doing a scatter plot because it is the best representation of the data, especially since we are adding a regression line. Since there was an obvious extreme value, I set the x limit to 20, so the x axis only goes to the value of 20. I chose to go to 20 because I think it better represents the entireity of the data, and provides a regression line that better represents the trend. I chose for the y axis to go to 100 to get rid of extreme values. The xlab() and ylab() functions allow me to label the axes. geom_smooth() creates the regression line (with the method = lm). I use "se=F" to get rid of the gray confidence interval. Lastly, I add my theme to the plot, and print it, so I can look at the outcome.  


```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and 
(c) TN, with month as the x axis and lake as a color aesthetic. Then, create 
a cowplot that combines the three graphs. Make sure that only one legend is 
present and that graph axes are aligned. 

Tip: R has a build in variable called `month.abb` that returns a list of months;
see <https://r-lang.com/month-abb-in-r-with-example>

```{r}
#5

Temperature_Month <- ggplot(Chemistry_Nutrients_PeterPaul, aes(x =  factor(month, levels = c(1:12)), y = temperature_C, color = lakename)) +
  geom_boxplot() + 
  scale_x_discrete(labels = month.abb, drop= FALSE) +
  xlab("Month") + 
  ylab("Temperature °C") + 
  labs(color = "Lake Name") +
  mytheme
print(Temperature_Month)

#description...

TP_Month <- ggplot(Chemistry_Nutrients_PeterPaul, aes(x = factor(month, levels = c(1:12)), y = tp_ug, color = lakename)) +
  geom_boxplot() + 
  scale_x_discrete(labels = month.abb, drop= FALSE) +
  xlab("Month") + 
  ylab("Total Phosphorus (µg)") + 
  labs(color = "Lake Name") +
  mytheme
print(TP_Month)


#description...

TN_Month <- ggplot(Chemistry_Nutrients_PeterPaul, aes(x = factor(month, levels = c(1:12)), y = tn_ug, color = lakename)) +
  geom_boxplot() + 
  scale_x_discrete(labels = month.abb, drop= FALSE) +
  xlab("Month") + 
  ylab("Total Nitrogen (µg)") + 
  labs(color = "Lake Name") +
  mytheme
print(TN_Month)


#description...


Chem_Nutrient_Plot_Grid <- plot_grid(
  Temperature_Month + theme(legend.position = "none"),
  TP_Month + theme(legend.position = "none"),
  TN_Month + theme(legend.position = "bottom"),
 labels = c("Temperature", "Phosphorus", "Nitrogen"),
  nrow = 3, ncol = 1, align = "v", rel_heights = c(2,2,2.5))
print(Chem_Nutrient_Plot_Grid)


#description...

```

Question: What do you observe about the variables of interest over seasons 
and between lakes? 

> Answer: 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the 
"Needles" functional group. Plot the dry mass of needle litter by date and 
separate by NLCD class with a color aesthetic. (no need to adjust the name 
of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into 
three facets rather than separated by color. 

```{r}
#6 
drymass_date <- ggplot(subset(NiwotRidge_Litter, functionalGroup == "Needles"), aes(x = collectDate, y = dryMass, color = nlcdClass)) +
  geom_point() +
  xlab("Date Collected") + 
  ylab("Dry Mass of Needles (g)") +
  labs(color = "NLCD Class") +
  mytheme
print(drymass_date)

# To create this plot first, I used the ggplot() function to define the data I was plotting. I use the subset() function to ensure that I am only plotting points that are in the functional group of needles. After I get the subset, I am able to define the x and y axis as the date and the dry mass respectively. Using the color = nlcdClass allows me to assign a color to the points representing each of the nlcd classes. The geom_point() allows me to graph the scatter plot. The xlab() and ylab() allow me to name the axes respectively. The labs() function is similar, but this allows me to name the legend. 

#7
drymass_date_facet <- ggplot(subset(NiwotRidge_Litter, functionalGroup == "Needles"), aes(x = collectDate, y = dryMass)) +
  geom_point() +
  facet_wrap(vars(nlcdClass)) +
    ylab("Dry Mass of Needles (g)") + 
    xlab("Date Collected") +
  mytheme
print(drymass_date_facet)

# To create this plot I used the same subset as above. The only difference to this plot is the facet_wrap() function which allows me to separate the plots by the different nlcd classes. 


```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: I think plot 7 is more effective because you can more effectively see the different 